Real-Time Estimation of COVID-19 Infections via Deconvolution and Sensor Fusion

### 2022-02-24 paper/paper.tex
Reproduction -> reproductive
Correct typo, copyedit

The latent setting exhibits another degree of diversity within itself. In our
work, we target symptomatic COVID-19 infections, which, to be perfectly clear,
is a latent time series. Another example along similar lines is
\citet{Goldstein:2009}, who estimate influenza infection incidence via Bayesian
deconvolution of mortality data. Meanwhile, other authors might view inferring 
latent infections as just a stepping stone toward ultimately estimating the 
instantaneous reproductive number $R_t$, a key epidemic parameter. Important 
contributions to the methodology on real-time estimation of $R_t$ include: 
%\citet{Wallinga:2004}, who propose a forward-looking approach (in fact, to
%estimate a related but slightly different quantity known as the case
%reproduction number) based on an estimate of the serial interval;
\citet{Bettencourt:2008}, who use a local approximation to the SIR model, and 
\citet{Cori:2013, Thompson:2019}, who use a discretization of the renewal
equation within a Bayesian framework. For a thorough review and comparison of
these methods, see \citet{Gostic:2020}. The latter paper also discusses in some
detail the importance of properly modeling the delay between infection onset and
case report, and the issue of right truncation, which, as we will see, are
central issues in our paper as well.

@@

For location $\ell$, let $y_{\ell,t}$ denote the number of new cases reported at
$t$ and the number of new infections that have onset at time $t$, respectively,
per 100,000 people. Note that $y_{\ell,t},x_{\ell,t}$ obey the
model~\eqref{eq:conv_model2}, \eqref{eq:delay_prob2}, as we have just rescaled
the underlying counts here by a constant (to put them on the scale of rates),
and recall, we assume all locations have the same delay distribution
(Assumption~\ref{asm:geo_invar}).

@@

The instantaneous reproductive number $R_t$, the average number of secondary
infections at time $t$ generated from a primary infection in the past, is a
useful and interpretable parameter that reflects the dynamics of epidemic growth
in a population. In the SIR model, the instantaneous reproduction number $R_t$
and growth rate $r_t$ at time $t$ obey the following relationship:
\[
R_t \approx 1 + \frac{r_t}{\gamma}, 
\] 
where $\gamma$ denotes the recovery rate in the SIR model. While this is
well-known in the literature on mathematical modeling of epidemics (and is exact
under local exponential growth; see, e.g., \citet{Wallinga:2007}), its use in
the presence of confounding seems to be underexplored and potentially
undervalued. If $I_t$ denotes the number of new infections at $t$, then using a
simple discrete difference approximation to $r_t$ leads to: 
\[
R_t \approx 1 + \frac{1}{\gamma}\bigg( \frac{I_{t+1}}{I_t} - 1 \bigg). 
\]
A similar though not identical approximation is given in
\citet{Bettencourt:2008}, where $I_{t+1}/I_t-1$ is replaced by
$\log(I_{t+1}/I_t)$. Critically, incident infections only enter right-hand side
above as a \emph{ratio} of values adjacent in time, and thus if we are only able
to estimate this up to an unknown multiplicative factor (due to confounding),
then this factor approximately cancels in the ratio as long as it is slowly
varying in time. In slightly more detail (and for simplicity, considering just a
single location), suppose as before that a fraction $a_t$ of infections go
untested at time $t$. Then $I_t = b_t x_t$ where $x_t$ is the number of new
infections at time $t$ that show up in case reports (i.e., the focus of this
paper) and $b_t = 1/(1-a_t)$. From the previous display,   
\[
R_t \approx 1 + \frac{1}{\gamma}\bigg( \frac{b_{t+1} x_{t+1}}{b_t x_t} - 1
\bigg) \approx 1 + \frac{1}{\gamma}\bigg( \frac{x_{t+1}}{x_t} - 1 \bigg),
\]
where the last approximation is motivated by an additional assumption the 
untested fraction varies slowly over time (so $b_{t+1}/b_t \approx
1$). This shows that estimates of $x_t$ can produce \emph{approximately
  unconfounded} estimates of $R_t$, even though $x_t$ is itself confounded
due to a lack of universal testing. This is true both in the retrospective and 
real-time sense, and will be the topic of future study.

@@

% The instantaneous reproduction number in this model is defined as $R_t = 
% S(t) \beta / \gamma$. We can interpret the $R_t$ as being determined by 
% the slope of a natural first-order approximation to the curve $\log I(t)$ at
% time $t$. That is, looking back at the differential equation for $I(t)$, if we
% forget about the fact that $S(t)$ actually depends on $t$, this ``looks like'' 
% $I'(t)$ is a ``constant'' times $I(t)$ (that ``constant'' being $\beta S(t) -
% \gamma$), which suggests that for $t$ near a point $t_0$,   
% \begin{align}
% \nonumber
% I(t) &\approx I(t_0) \exp\big( (\beta S(t) - \gamma) (t-t_0) \big) \\ 
% \label{eq:inf_approx}
% &= I(t_0) \exp\big(\gamma (R_t - 1) (t-t_0)\big).
% \end{align}

>>>>

The latent setting exhibits another degree of diversity within itself. In our
work, we target symptomatic COVID-19 infections, which, to be perfectly clear,
is a latent time series. Another example along similar lines is
\citet{Goldstein:2009}, who estimate influenza infection incidence via Bayesian
deconvolution of mortality data. Meanwhile, other authors might view inferring 
latent infections as just a stepping stone toward ultimately estimating the 
instantaneous reproductive number $R_t$, a key epidemic parameter. Important 
contributions to the methodology on real-time estimation of $R_t$ include: 
%\citet{Wallinga:2004}, who propose a forward-looking approach (in fact, to
%estimate a related but slightly different quantity known as the case
%reproductive number) based on an estimate of the serial interval;
\citet{Bettencourt:2008}, who use a local approximation to the SIR model, and 
\citet{Cori:2013, Thompson:2019}, who use a discretization of the renewal
equation within a Bayesian framework. For a thorough review and comparison of
these methods, see \citet{Gostic:2020}. The latter paper also discusses in some
detail the importance of properly modeling the delay between infection onset and
case report, and the issue of right truncation, which, as we will see, are
central issues in our paper as well.

@@

For location $\ell$, denote by $y_{\ell,t}$ and $x_{\ell,t}$ the number of new
cases reported and number of new infections that onset at time $t$,
respectively, per 100,000 people. Note that $y_{\ell,t},x_{\ell,t}$ obey
\eqref{eq:conv_model2}, \eqref{eq:delay_prob2}, because we have just rescaled
the underlying counts here by a constant (in order to put them on the scale of
rates), and recall, we assume that all locations have the same delay
distribution (Assumption~\ref{asm:geo_invar}).

@@

The instantaneous reproductive number $R_t$, the average number of secondary
infections at time $t$ generated from a primary infection in the past, is a
useful and interpretable parameter that reflects the dynamics of epidemic growth
in a population. In the SIR model, the instantaneous reproductive number $R_t$
and growth rate $r_t$ at time $t$ obey the following relationship:
\[
R_t \approx 1 + \frac{r_t}{\gamma}, 
\] 
where $\gamma$ denotes the recovery rate in the SIR model. While this is
well-known in the literature on mathematical modeling of epidemics (and is exact
under local exponential growth; see, e.g., \citet{Wallinga:2007}), its use in
the presence of confounding seems to be underexplored and potentially
undervalued. If $I_t$ denotes the number of new infections at $t$, then using a
simple discrete difference approximation to $r_t$ leads to: 
\[
R_t \approx 1 + \frac{1}{\gamma}\bigg( \frac{I_{t+1}}{I_t} - 1 \bigg). 
\]
A similar though not identical approximation is given in
\citet{Bettencourt:2008}, where $I_{t+1}/I_t-1$ is replaced by
$\log(I_{t+1}/I_t)$. Critically, incident infections only enter right-hand side
above as a \emph{ratio} of values adjacent in time, and thus if we are only able
to estimate this up to an unknown multiplicative factor (due to confounding),
then this factor approximately cancels in the ratio as long as it is slowly
varying in time. In slightly more detail (and for simplicity, considering just a
single location), suppose as before that a fraction $a_t$ of infections go
untested at time $t$. Then $I_t = b_t x_t$ where $x_t$ is the number of new
infections at time $t$ that show up in case reports (i.e., the focus of this
paper) and $b_t = 1/(1-a_t)$. From the previous display,   
\[
R_t \approx 1 + \frac{1}{\gamma}\bigg( \frac{b_{t+1} x_{t+1}}{b_t x_t} - 1
\bigg) \approx 1 + \frac{1}{\gamma}\bigg( \frac{x_{t+1}}{x_t} - 1 \bigg),
\]
where the last approximation is motivated by an additional assumption the 
untested fraction varies slowly over time (so $b_{t+1}/b_t \approx
1$). This shows that estimates of $x_t$ can produce \emph{approximately
  unconfounded} estimates of $R_t$, even though $x_t$ is itself confounded
due to a lack of universal testing. This is true both in the retrospective and 
real-time sense, and will be the topic of future study.

@@

% The instantaneous reproductive number in this model is defined as $R_t = 
% S(t) \beta / \gamma$. We can interpret the $R_t$ as being determined by 
% the slope of a natural first-order approximation to the curve $\log I(t)$ at
% time $t$. That is, looking back at the differential equation for $I(t)$, if we
% forget about the fact that $S(t)$ actually depends on $t$, this ``looks like'' 
% $I'(t)$ is a ``constant'' times $I(t)$ (that ``constant'' being $\beta S(t) -
% \gamma$), which suggests that for $t$ near a point $t_0$,   
% \begin{align}
% \nonumber
% I(t) &\approx I(t_0) \exp\big( (\beta S(t) - \gamma) (t-t_0) \big) \\ 
% \label{eq:inf_approx}
% &= I(t_0) \exp\big(\gamma (R_t - 1) (t-t_0)\big).
% \end{align}
